\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-11-01 paper/paper.tex
Last experiments.

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--30 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|cc|cc|cc}
\hline
 & & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} & \multicolumn{2}{c}{\textbf{Read Server}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &      -- \\
     & NA12891 &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &      -- \\
     & NA12892 &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &      -- \\
     & Merged  &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &      -- \\
\hline
\RS  & AA      &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB & 48.3 GB \\
     & TT      &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB & 48.4 GB \\
     & AT      &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB & 32.6 GB \\
     & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB & 40.6 GB \\
     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB &  152 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &  281 GB \\
     & Merged  & 2.45 Tbp & 26.5G &      -- &         &         &         \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &  281 GB \\
     & Merged  & 2.44 Tbp & 26.5G &      -- &         &         &         \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

For the further experiments, we chose 128~MB run buffers, and 256~MB thread buffers and 6 merge buffers (overhead 30.8~GB).

@@

The results can be seen in Figure~\ref{fig:benchmark} (right).
% FIXME discussion of the results; note that ropebwt2 did not finish without RLO; BWT-merge does not do RLO

\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files.
% FIXME sizes in native/sga format, merging time/space

@@

We have proposed an improved \BWT{} merging algorithm for large read collections. Our implementation of the algorithm in the \BWTmerge{} tool is fast enough to be used with terabases of sequence data. It requires only 20\nobreakdash--30~gigabytes of memory on top of the \BWT{}s to be merged. As \BWT-based indexes do random accesses to large arrays, they must reside in memory in most applications. Hence \BWTmerge{} can build the index on the same system as it is going to be used.

>>>>

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. With our largest 2.4 Tbp datasets, the algorithm is capable of merging 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format. RLO indicates that the reads are sorted in reverse lexicographic order. The numbers in parentheses are estimates.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|cc|cc|cc}
\hline
 & & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} & \multicolumn{2}{c}{\textbf{Read Server}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &       -- \\
     & NA12891 &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &       -- \\
     & NA12892 &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &       -- \\
     & Merged  &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &       -- \\
\hline
\RS  & AA      &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB &  48.3 GB \\
     & TT      &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB &  48.4 GB \\
     & AT      &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB &  32.6 GB \\
     & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB & (152 GB) \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &   281 GB \\
     & Merged  & 2.45 Tbp & 26.5G &      -- &  181 GB &  197 GB & (239 GB) \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &   281 GB \\
     & Merged  & 2.44 Tbp & 26.5G &      -- &  180 GB &  197 GB & (238 GB) \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).

@@

The results can be seen in Figure~\ref{fig:benchmark} (right). When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt. \ropebwt2{} ran out of memory when processing the last 10~Gbpp batch of reads. \BWTmerge{} would have been about 1.2x slower and 1.5x more space-efficient than it. The running time of \BWTmerge{} was split almost evenly between building the \BWT{}s of individual files and merging them.

When we tell \ropebwt{} and \ropebwt2{} to sort the reads in reverse lexicographic order, all tools improve their performance. \BWTmerge{} is now 1.70x slower and 2.12x more space-efficient than \ropebwt{} and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2. Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them. Note that the final \BWT{} built by \BWTmerge{} is not in reverse lexicographic order, unlike the ones built with \ropebwt{} and \ropebwt2{}. When such functionality is implemented, \BWTmerge{} should be somewhat slower and somewhat more space-efficient than when maintaining the existing order of reads.

\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC in the first collection; TT, GT, CT, AT, TG, GG, CG, AG in the second one). Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~gigabytes and 219~gigabytes of memory, and used 297~gigabytes and 300~gigabytes of disk space, respectively. We reduced the size of the FM-indexes from around 560~GB to 480~GB in the format used in the Read Server. With the native format of \BWTmerge{}, the total size of the indexes was down to 360~GB. This will make it possible to host the indexes on two servers instead of the original three.

@@

We have proposed an improved \BWT{} merging algorithm for large read collections. Our implementation of the algorithm in the \BWTmerge{} tool is fast enough to be used with terabases of sequence data. It requires only 30~gigabytes of memory on top of the \BWT{}s to be merged. As \BWT-based indexes do random accesses to large arrays, they must reside in memory in most applications. Hence \BWTmerge{} can build the index on the same system as it is going to be used.
